---
title: "Final data joining syntax"
author: "Mason Walker"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Joining all the datasets

Here I will joing the datasets we are using for our capstone project together to form our final dataset. The datasets that will stitched together include:

- U.S. Census ACS county demographic and housing data
- BLS county unenmployment data
- BLS county industry and wage concentration data
- Zillow county housing values index data

The respective data files can be found on GitHub and include:

- `acs_final.csv` = U.S. Census ACS county data
- `bls_final.csv` = BLS county unenemployment data
- `zillow_final.csv` = Zillw county housing values index data
- `bls_qcew_final.csv` = BLS county industry and wage concentration data

```{r joing data}
# Loading libraries
library(tidyverse)

# Loading all the data
census_data <- read.csv("acs_final.csv", colClasses = c('fips'='character'))
bls_employment_data <- read.csv("bls_final.csv", colClasses = c('fips'='character'))
zillow_data <- read.csv("zillow_final.csv", colClasses = c('fips'='character'))
bls_qcew_data <- read.csv("bls_qcew_final.csv", colClasses = c('area_code'='character'))


# Drop repetive variables

bls_employment_data <- bls_employment_data |>
  select(-c(X, state_name, county))

bls_qcew_data <- bls_qcew_data |>
  select(-c(X, area))

zillow_data <- zillow_data |>
  select(-c(X, region_id, size_rank, region_name, region_type, state_name, metro,
            state_code_fips, county_code_fips))

final_data <- census_data |>
  left_join(bls_employment_data, by = join_by(fips == fips)) |>
  left_join(bls_qcew_data, by = join_by(fips == area_code)) |>
  left_join(zillow_data, by = join_by(fips == fips)) |>
  select(-c(X))

# Outputting the final data
write.csv(final_data, "final_data.csv")



```

